---
title: "Final Project Report - Practical Machine Learning Course"
author: "Ruben Maso"
date: "24 de diciembre de 2015"
output:
  html_document:
    keep_md: yes
---

# Background Introduction
These are the files produced during a homework assignment of Coursera's MOOC Practical Machine Learning from Johns Hopkins University. Here is the introduction of the exercise:

"Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement ??? a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: [http://groupware.les.inf.puc-rio.br/har](see the section on the Weight Lifting Exercise Dataset)."


Data Sources
The training data for this project are available here:

[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv]

The test data are available here:

[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv]

The data for this project comes from this original source: [http://groupware.les.inf.puc-rio.br/har]. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

Please Note that I the code I use loads the data from the file system, so that you are must download the files to your environment. Please customize the code to your specific needs.

# Project Intended Results
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).

2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.

```{r warning = FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rpart.plot)
library(randomForest)
```


Finally, load the same seed with the following line of code:
```{r}
set.seed(666)
```

# Getting the data
The training and testing data files:

```{r}
trainUrl <- "pml-training.csv"
testUrl <- "pml-testing.csv"
```

Load data 
```{r}
training <- read.csv(trainUrl, na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(testUrl, na.strings=c("NA","#DIV/0!",""))
```

# Partioning data
Partioning Training data set into two data sets, 60% for myTraining, 40% for myTesting:

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]; myTesting <- training[-inTrain, ]
dim(myTraining); dim(myTesting)
```

# Cleaning the data
The following transformations were used to clean the data:

### Transformation 1: Cleaning NearZeroVariance Variables 
Run this code to view possible NZV Variables:

```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
```

Run this code to create another subset without NZV variables:

```{r}
myNZVvars <- names(myTraining)[!myDataNZV$nzv]
myTraining <- myTraining[myNZVvars]
dim(myTraining)
```

Transformation 2: Killing first column of Dataset - ID Removing first ID variable so that it does not interfer with ML Algorithms:

```{r}
myTraining <- myTraining[c(-1)]
```

Transformation 3: Cleaning Variables with too many NAs. For Variables that have more than a 60% threshold of NA's I'm going to leave them out:

```{r}
trainingV3 <- myTraining #creating another subset to iterate in loop
for(i in 1:length(myTraining)) { #for every column in the training dataset
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if n?? NAs > 60% of total observations
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
                trainingV3 <- trainingV3[ , -j] #Remove that column
            }   
        } 
    }
}
#To check the new N?? of observations
dim(trainingV3)
#Seting back to our set:
myTraining <- trainingV3
rm(trainingV3)
```


Now let us do the exact same 3 transformations but for our myTesting and testing data sets.
```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -length(myTraining)]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2]

#To check the new N?? of observations
dim(myTesting)
#To check the new N?? of observations
dim(testing)
#Note: The last column - problem_id - which is not equal to training sets, was also "automagically" removed
#No need for this code:
#testing <- testing[-length(testing)]
```

# Using ML algorithms for prediction: Decision Tree
```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
```

To view the decision tree with fancy run this command:
```{r}
rpart.plot(modFitA1)
# fancyRpartPlot(modFitA1)
```

Predicting:

```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```

Using confusion Matrix to test results:
```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```


# Using ML algorithms for prediction: Random Forests
```{r}
fitControl <- trainControl(## 10-fold CV
                           method = "cv",
                           # Reduce number to perform faster
                           number = 3
                           )
rfFit <- train(classe ~ ., data = myTraining,
                 method = "rf",
                 trControl = fitControl,
                 verbose = TRUE)
```

Predicting in-sample error:
```{r}
rfPredictions <- predict(rfFit, myTesting, type = "raw")
```

Using confusion Matrix to test results:
```{r}
confusionMatrix(rfPredictions, myTesting$classe)
```


# Using ML algorithms for prediction: GBM
```{r}
# gbmGrid <-  expand.grid(interaction.depth = c(1, 5, 9),
#                         n.trees = (1:30)*50,
#                         shrinkage = 0.1,
#                         n.minobsinnode = 20)
gbmFit <- train(classe ~ ., data = myTraining,
                 method = "gbm",
                 trControl = fitControl,
                 ## This last option is actually one
                 ## for gbm() that passes through
                 verbose = TRUE)
                 ## Now specify the exact models 
                 ## to evaluate:
                 # tuneGrid = gbmGrid)
gbmFit
```

Predicting in-sample error:
```{r}
gbmPredictions <- predict(gbmFit, myTesting, type = "raw")
```

Using confusion Matrix to test results:
```{r}
confusionMatrix(gbmPredictions, myTesting$classe)
```



# Using ML algorithms for prediction: LogitBoost
```{r}
# gbmGrid <-  expand.grid(interaction.depth = c(1, 5, 9),
#                         n.trees = (1:30)*50,
#                         shrinkage = 0.1,
#                         n.minobsinnode = 20)
logitBoostFit <- train(classe ~ ., data = myTraining,
                 method = "LogitBoost",
                 trControl = fitControl,
                 verbose = TRUE)
logitBoostFit
```

Predicting in-sample error:
```{r}
logitBoostPredictions <- predict(logitBoostFit, myTesting, type = "raw")
```

Using confusion Matrix to test results:
```{r}
confusionMatrix(logitBoostPredictions, myTesting$classe)
```

# Generating Files to submit as answers for the Assignment:
Finally, using the provided Test Set out-of-sample error.

For Random Forests we use the following formula, which yielded a much better prediction in in-sample:

```{r}
predictionsB2 <- predict(rfFit, testing, type = "raw")
```

Function to generate files with predictions to submit for assignment
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictionsB2)
```

